Co-methylation

### 2015-04-23 Rmarkdown/comethylation.Rmd
Switch summation notation. Move code from thesis to supplementary material.

The possible values of $\psi$ are highly skewed, with $0 < \psi < 1$ and $1 < \psi < \infty$ corresponding to the two distinct dependence scenarios. Rather than estimating $\psi$ directly, we will instead focus our attention on the _log odds ratio_, $\phi = \log{\psi}$. The log odds ratio is symmetric about 0, which makes statistical inference somewhat simpler. I will routinely switch between discussing the estimation of $\psi$ and $\phi$, noting that we can always convert one to the other by appropriate exponentiation or taking of logarithms.

Suppose we observe $(X, Y)$ for $n$ samples. Denote by $n_{ij}, i, j = 1, 2$ the number of subjects with response $i$ for variable $X$ and response $j$ for variable $Y$. We use the 'dot' notation to denote the sum over the index it replaces, e.g., $n_{1 \cdot} = n_{11} + n_{12}$. The general form of this $2 \times 2$ contingency table is shown in Table \ref{tab:2x2}.

\begin{table}[h]
\centering
\begin{tabular}{@{}cccc@{}}
\cmidrule(l){2-4}
 & \multicolumn{3}{c}{Column} \\
 \cmidrule(l){2-4}
\multirow{3}{*}{Row} & & & \\
 & $n_{11}$       & $n_{12}$       & $n_{1 \cdot}$ \\
 & $n_{21}$       & $n_{22}$       & $n_{2 \cdot}$ \\
 & $n_{\cdot 1}$  & $n_{\cdot 2}$  & $n_{\cdot \cdot} = n$ \\ \cmidrule(l){2-4}
\end{tabular}
\caption{Notation for a $2 \times 2$ contingency table.}
\label{tab:2x2}
\end{table}

The simplest estimator of $\phi$ is the unconditional maximum likelihood estimator, $\widehat{\phi_{U}} = \log{\frac{n_{11} n_{22}}{n_{12} n_{21}}}$. The asymptotic standard error of $\widehat{\phi_{U}}$ is given by $\hat{\sigma}(\widehat{\phi_{U}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$. The asymptotic distribution of $\widehat{\phi_{U}}$ as $n \to \infty$ is $Normal(\phi, \hat{\sigma}(\phi)^{2})$.

A problem with the estimator $\widehat{\phi_{U}}$ is that it is $0$ or $\infty$ if any $n_{ij} = 0$ and undefined if either the row or column sums are zero, events that have positive probabilities. We can avoid such problems by adding $\frac{1}{2}$ to each of the $n_{ij}$ and defining a modified estimator $\widehat{\phi_{0.5}} = \log{\frac{(n_{11} + 0.5) (n_{22} + 0.5)}{(n_{12} + 0.5) (n_{21} + 0.5)}}$ with corresponding asymptotic standard error $hat{\sigma}(\widehat{\phi_{0.5}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$. \citet{Haldane:1956wq} and \citet{Anscombe:1956te} showed that $\widehat{\phi_{U}}$ and $\widehat{\phi_{0.5}}$ have the same asymptotic distribution around $\phi$ as $n \to \infty$ but that $\widehat{\phi_{0.5}}$ has reduced first-order bias. The modified estimator, $\widehat{\phi_{0.5}}$, is therefore generally recommended.

Another alternative to $\widehat{\phi_{U}}$ is the _conditional_ maximum likelihood estimator, $\widehat{\phi_{C}}$. This is computed by first conditioning on ${n_{1 \cdot}, n_{\cdot 1}}$. Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 \cdot}, n_{\cdot, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}. The maximum likelihood estimator of $\psi$, $\widehat{psi_{C}}$ is solved using iterative methods.

The conditional estimator, $\widehat{\phi_{C}}$, works better than the unconditional estimator, $\widehat{\phi_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}. Estimation of the odds ratio is made more difficult when the contingency table is _sparse_. We say that a contingency table is sparse when some of the $n_{ij}$ are small. Sparsity can occur even when the sample size, $n$, is large. Sparsity becomes more of a problem as we move from a single $2 \times 2$ table to $K$ $2 \times 2$ tables or, equivalently, a $2 \times 2 \times K$ table.

@@

Suppose we now measure the same two binary variables $X$ and $Y$ in $K$ different strata. We summarise these data in a $2 \times 2 \times K$ contingency table, $\{n_{ijk} \}$, where the $k^{th}$ 'slice' of the table corresponds to the $2 \times 2$ table for the $k^{th}$ stratum. For now, we assume that the odds ratio, $\psi = \frac{\pi_{11k} \pi_{22k}}{\pi_{12k} \pi_{21k}}$, remains constant across the $K$ tables but we allow the marginal probabilities, $\pi_{\cdot 1k} = 1 - \pi_{\cdot 2k}, \pi_{1 \cdot k} = 1 - \pi_{2 \cdot k}$ to vary.

We can again compute an unconditional maximum likelihood estimator, $\widehat{phi_{U}}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{phi_{0.5}}$, and a conditional maximum likelihood estimator, $\widehat{phi_{C}}$, of the log odds ratio $\phi = \log{\psi}$ \citep[see][]{Breslow:1981ta}.

Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi_{MH}} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{\cdot \cdot k})}{n_{12k} n_{21k} / n_{\cdot \cdot k}}$. \citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\phi_{MH}} = \log{\widehat{\psi_{MH}}}$, $\hat{\sigma}^{2}(\widehat{\phi_{MH}})$. This variance estimator did not appear until some 25 years after the initial publication of the Mantel-Haenszel estimator[^mh_variance]. This delay was in part due to the two asymptotic forms of $2 \times 2 \times K$ contingency tables.

@@

The first asymptotic environment, $A1$, assumes that the number of $2 \times 2$ tables, $K$, remains small while the the sample sizes in each strata, $n_{\cdot \cdot k}$, grows large. The second asymptotic environment, $A2$, assumes that $K$ grows while $n_{\cdot \cdot k}$ remains small. The variance estimator proposed by \citet{Robins:1986vh} is consistent under both $A1$ and $A2$. For our study of within-fragment co-methylation, $A2$ is the more relevant asymptotic environment.

@@

Under asymptotic environment $A2$, where the $2 \times 2$ tables are sparse, \citet{Liang:1985uw} developed three tests for the assumption of homogeneity of odds ratios and compared these with two tests for the same assumption but designed under $A1$. Unsurprisingly, \citeauthor{Liang:1985uw} found that the three conditional tests are more appropriate than either of the unconditional tests when the data are simulated under $A2$. The disadvantage of these conditional tests is the extra computation required to compute them. \citeauthor{Liang:1985uw} also note, however, that when both $K > 10$ and $n_{i \cdot} > 10$, the simplest test, an unconditional score test developed under $A1$ by \citet{Zelen:1971fo} and corrected by \citet{Halperin:1977vk}, has approximately the correct size at the nominal 0.05 level.

@@

- the number of tables, $K$. The number of 2-tuples.
- the marginal probabilities, $\pi_{1 \cdot} = Pr(Z_{i})$ and $\pi_{\cdot 1} = Pr(Z_{i'})$. Some simulations use a common set of marginal probabilities across all tables but some simulate the marginal probabilities from a specified distribution.
- the odds ratio of each table, $\psi_{k}$. Most simulations use a common odds ratio across the $K$ tables, $\psi_{k} = \psi$.
- the sample size of each table, $n_{\cdot \cdot k} = d_{(i, i')}$. The number of reads containing both the $i^{th}$ and ${i'}^{th}$ methylation locus (sequencing depth). Most simulations use a truncated negative binomial distribution to model a bisulfite-sequencing experiment with mean sequencing depth of $30 \times$ but where only 2-tuples with at least $10 \times$ sequencing depth are used in downstream analyses[^nb_sim].

@@

- Use parameters $n, \pi, \psi, \phi$ and not the ones from Chapter \ref{chap:wgbs_statistical_framework} since these are more succinct.

@@

We will be studying within-fragment co-methylation by analysing a large number of sparse $2 \times 2$ contingency tables. This puts us in the realm of asymptotic environment $A2$.
Furthermore, the marginal probabilities of these tables, $\pi_{\cdot 1} = 1 - \pi_{\cdot 2}$ and $\pi_{1 \cdot} = 1 - \pi_{2 \cdot}$, will be very often close to the boundaries of zero and one. This is because in this case the marginal probability represents the probability that the methylation locus is methylated, and we have seen in Chapter \ref{chap:wgbs_statistical_framework} that these values, as estimated by $\beta$-values, have a bimodal distribution with each mode near the boundary.

@@

__TODO: Move this code chunk to supplementary material and only include results in thesis.__

```{r}
# Requires the installation of methsim
library(mipfp)

# rtpois is a helper function to simulate n truncated Poisson(lambda) RVs with
# truncation specified by 'trunc'.
# It is adapted from a method for simulating zero-truncated Poisson suggested
# by Peter Dalgaard https://stat.ethz.ch/pipermail/r-help/2005-May/070680.html
rtpois <- function(n, lambda, trunc) {
  qpois(runif(n, ppois(trunc, lambda), 1), lambda)
}

# rtnbinom is a helper function to simulate n truncated NB(mu, size) RVs with
# truncation specified by 'trunc'. It uses a more explicit approach to simulate # truncated RVs than rtpois() because the same trick doesn't immediately work
# for NB-distributed RVs.
rtnbinom <- function(n, mu, size, trunc) {
  x <- rnbinom(n, size = size, mu = mu)
  i <- x <= trunc
  while (any(i)) {
    x[i] <- rnbinom(sum(i), size = size, mu = mu)
    i <- x <= trunc
  }
  x
}

# Function to simulate 2x2xK tables.
# Each 2x2 table represents the counts of unmethylated (0) and methylated (1) states
# at each CpG for each read.
# The number of reads per table (i.e., 2x2 table sample size) is simulated
# from a truncated Poisson(lambda) distribution; truncation is to simulate
# minimum coverage cutoffs and is specified by the 'trunc' argument.
# The marginal probabilities of each table are p1 = Prob(methylated at first CpG)
# and p2 = Prob(methylated at second CpG).
# The odds ratio (or) is for each 2x2 table and should be of length K or 1 (in
# which case all K tables have the same odds ratio).
simTables <- function(K, lambda, trunc, p1, p2, or) {

  # d is the sample size of each 2x2 table (i.e, the sequencing depth for the 2-tuple)
  d <- rtpois(n = K, lambda = lambda, trunc = trunc)
  if (length(p1) == 1L) {
    p1 <- rep(p1, K)
  }
  if (length(p2) == 1L) {
    p2 <- rep(p2, K)
  }
  if (length(or) == 1L) {
    or <- rep(or, K)
  }
  .simTables <- function(k, d, p1, p2, or) {
    # Compute joint probabilities
    P <- mipfp::Ipfp(seed = matrix(c(or, 1, 1, 1), ncol = 2),
                     target.list = list(1, 2),
                     target.data = list(c(1 - p1, p1), c(1 - p2, p2)))$x.hat
    matrix(rmultinom(1, d, as.vector(P)), ncol = 2)
  }
  mapply(.simTables, k = K, d = d, p1 = p1, p2 = p2, or = or, SIMPLIFY = "array")
}

# Estimator of the (offset-adjusted) unconditional log odds ratios of K 2x2 tables
LOR <- function(x, offset = 0.5) {
  x <- x + offset
  MM <- x[2, 2, ]
  MU <- x[2, 1, ]
  UM <- x[1, 2, ]
  UU <- x[1, 1, ]
  log2(MM) + log2(UU) - log2(MU) - log2(UM)
}

# Mantel-Haenszel estimator of a common log odds ratio of a 2x2xK table
MH <- function(x, offset = 0) {
  x <- x + offset
  # TODO: return the standard error
  MM <- x[2, 2, ]
  MU <- x[2, 1, ]
  UM <- x[1, 2, ]
  UU <- x[1, 1, ]
  d <- apply(x, 3, sum)
  or <- sum((UU * MM) / d) / sum((MU * UM) / d)
  # One or more margins must be identically zero so add offset
  if (is.nan(or) | isTRUE(all.equal(or, 0)) | is.infinite(or)) {
    warning("One or more cell is (probably) identically zero in all K 2x2 tables. Adding offset = 0.5")
    MH(x, 0.5)
  } else {
    log2(or)
  }
}

# Estimator of the (offset-adjusted) unconditional log odds ratio of the
# 2x2xK collapsed over K.
LORCollapsed <- function(x, offset = 0.5) {
  z <- apply(x, c(1, 2), sum)
  # Artificially augment so it's a 2x2x1 table
  dim(z) <- c(dim(z), 1)
  lor <- LOR(z, 0)
  if (is.nan(lor) | is.infinite(lor)) {
    warning("One or more cell is (probably) identically zero in all K 2x2 tables. Adding offset = 0.5")
    LOR(z, 0.5)
  } else {
    lor
  }
}

# Simulate and plot data when p1 = p2.
# 100 CpG pairs, 30x average coverage truncated at 10
# odds ratio = 1, i.e., independence
set.seed(666)
par(mfrow = c(3, 3))
K <- 100
lambda <- 20
trunc <- 11
or <- 1
p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
p2 <- p1
for (i in 1:9) {
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1[i], p2 = p2[i], or = or)
  hist(LOR(x), main = paste0("p1 = ", p1[i], ", p2 = ", p2[i], ", lor = ", log2(or)),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  abline(v = log2(or), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# Simulate and plot data when p1 = p2.
# 1000 CpG pairs, 30x average coverage truncated at 10
# odds ratio = 1, i.e., independence
set.seed(666)
par(mfrow = c(3, 3))
K <- 1000
lambda <- 20
trunc <- 10
or <- 1
p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
p2 <- p1
for (i in 1:9) {
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1[i],
                 p2 = p2[i], or = or)
  hist(LOR(x),
       main = paste0("p1 = ", p1[i], ", p2 = ", p2[i], ", lor = ", log2(or)),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  abline(v = log2(or), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# Simulate and plot data when p1, p2 ~ Uniform(0, 1)
# 100 CpG pairs, 30x average coverage truncated at 10
# log odds ratio = -4, -3, ..., 3, 4.
set.seed(666)
par(mfrow = c(3, 3))
K <- 100
lambda <- 20
trunc <- 11
or <- 2^seq(-4, 4, 1)
p1 <- runif(K)
p2 <- runif(K)
for (i in 1:9) {
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2,
                 or = or[i])
  hist(LOR(x),
       main = paste0("p1 = U(0, 1), p2 = U(0, 1)", ", lor = ", log2(or[i])),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  abline(v = log2(or[i]), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc),
      outer = TRUE, line = -1)

# Simulate and plot data when p1 ~ Uniform(0, 1), p2 = p1
# 100 CpG pairs, 30x average coverage truncated at 10
# log odds ratio = -4, -3, ..., 3, 4.
set.seed(666)
par(mfrow = c(3, 3))
K <- 100
lambda <- 20
trunc <- 11
or <- 2^seq(-4, 4, 1)
p1 <- runif(K)
p2 <- p1
for (i in 1:9) {
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or[i])
  hist(LOR(x), main = paste0("p1 = U(0, 1), p2 = p1", ", lor = ", log2(or[i])),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       frequeny = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  abline(v = log2(or[i]), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc),
      outer = TRUE, line = -1)

# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
# 100 CpG pairs, 30x average coverage truncated at 10
# log odds ratio = -4, -3, ..., 3, 4.
# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation
# levels (B).
set.seed(666)
par(mfrow = c(3, 3))
K <- 100
lambda <- 20
trunc <- 11
or <- 2^seq(-4, 4, 1)
p1 <- rbeta(K, 0.3, 0.2)
p2 <- p1
for (i in 1:9) {
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or[i])
  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = ", log2(or[i])),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  abline(v = log2(or[i]), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
# 100 CpG pairs, 30x average coverage truncated at 10
# log2(or) ~ Gaussian(log2(or_mean), 1)
# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation
# levels (B).
set.seed(666)
par(mfrow = c(3, 3))
K <- 100
lambda <- 20
trunc <- 11
or_mean <- 2 ^ seq(-4, 4, 1)
p1 <- rbeta(K, 0.3, 0.2)
p2 <- p1
for (i in 1:9) {
  or <- 2 ^ rnorm(K, log2(or_mean[i]), 1)
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = Norm(", log2(or_mean[i]), ", 1)"),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  lines(density(log2(or)), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
# 1000 CpG pairs, 30x average coverage truncated at 10
# log2(or) ~ Gaussian(log2(or_mean), 1)
# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation
# levels (B).
set.seed(666)
par(mfrow = c(3, 3))
K <- 1000
lambda <- 20
trunc <- 11
or_mean <- 2 ^ seq(-4, 4, 1)
p1 <- rbeta(K, 0.3, 0.2)
p2 <- p1
for (i in 1:9) {
  or <- 2 ^ rnorm(K, log2(or_mean[i]), 1)
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = Norm(", log2(or_mean[i]), ", 1)"),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  lines(density(log2(or)), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
# 1000 CpG pairs, 30x average coverage truncated at 10
# log2(or) ~ Skewed-Gaussian(log2(or_mean), 1, -10)
# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation
# levels (B).
library(sn)
set.seed(666)
par(mfrow = c(3, 3))
K <- 1000
lambda <- 20
trunc <- 11
or_mean <- 2 ^ seq(-4, 4, 1)
p1 <- rbeta(K, 0.3, 0.2)
p2 <- p1
for (i in 1:9) {
  or <- 2 ^ rsn(K, log2(or_mean[i]), 1, -10)
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = SN(", log2(or_mean[i]), ", 1, -10)"),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  lines(density(log2(or)), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
# 1000 CpG pairs, 30x average coverage truncated at 10
# log2(or) ~ 0.5 * Gaussian(log2(or_mean), 1) + 0.5 * Gaussian(-log2(or_mean))
# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation
# levels (B).
set.seed(666)
par(mfrow = c(3, 3))
K <- 1000
lambda <- 20
trunc <- 11
or_mean <- 2 ^ seq(-4, 4, 1)
w <- sample(c(-1, 1), K, replace = TRUE)
p1 <- rbeta(K, 0.3, 0.2)
p2 <- p1
for (i in 1:9) {
  or <- 2 ^ rnorm(K, w * log2(or_mean[i]), 1)
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = NM(", log2(or_mean[i]), ", 1)"),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  lines(density(log2(or)), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# TODO: Increase d for each table
```

>>>>

The possible values of $\psi$ are highly skewed, with $0 < \psi < 1$ and $1 < \psi < \infty$ corresponding to the two distinct dependence scenarios. Rather than estimating $\psi$ directly, we will instead focus our attention on the _log odds ratio_, $\theta = \log{\psi}$. The log odds ratio is symmetric about 0, which makes statistical inference somewhat simpler. I will routinely switch between discussing the estimation of $\psi$ and $\theta$, noting that we can always convert one to the other by appropriate exponentiation or taking of logarithms.

Suppose we observe $(X, Y)$ for $n$ samples. Denote by $n_{ij}, i, j = 1, 2$ the number of subjects with response $i$ for variable $X$ and response $j$ for variable $Y$. We use the 'dot' notation to denote the sum over the index it replaces, e.g., $n_{1 +} = n_{11} + n_{12}$. The general form of this $2 \times 2$ contingency table is shown in Table \ref{tab:2x2}.

\begin{table}[h]
\centering
\begin{tabular}{@{}cccc@{}}
\cmidrule(l){2-4}
 & \multicolumn{3}{c}{Column} \\
 \cmidrule(l){2-4}
\multirow{3}{*}{Row} & & & \\
 & $n_{11}$       & $n_{12}$       & $n_{1 +}$ \\
 & $n_{21}$       & $n_{22}$       & $n_{2 +}$ \\
 & $n_{+ 1}$  & $n_{+ 2}$  & $n_{+ +} = n$ \\ \cmidrule(l){2-4}
\end{tabular}
\caption{Notation for a $2 \times 2$ contingency table.}
\label{tab:2x2}
\end{table}

The simplest estimator of $\theta$ is the unconditional maximum likelihood estimator, $\widehat{\theta_{U}} = \log{\frac{n_{11} n_{22}}{n_{12} n_{21}}}$. The asymptotic standard error of $\widehat{\theta_{U}}$ is given by $\hat{\sigma}(\widehat{\theta_{U}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$. The asymptotic distribution of $\widehat{\theta_{U}}$ as $n \to \infty$ is $Normal(\theta, \hat{\sigma}(\theta)^{2})$.

A problem with the estimator $\widehat{\theta_{U}}$ is that it is $0$ or $\infty$ if any $n_{ij} = 0$ and undefined if either the row or column sums are zero, events that have positive probabilities. We can avoid such problems by adding $\frac{1}{2}$ to each of the $n_{ij}$ and defining a modified estimator $\widehat{\theta_{0.5}} = \log{\frac{(n_{11} + 0.5) (n_{22} + 0.5)}{(n_{12} + 0.5) (n_{21} + 0.5)}}$ with corresponding asymptotic standard error $hat{\sigma}(\widehat{\theta_{0.5}}) = \sqrt{\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}}$. \citet{Haldane:1956wq} and \citet{Anscombe:1956te} showed that $\widehat{\theta_{U}}$ and $\widehat{\theta_{0.5}}$ have the same asymptotic distribution around $\theta$ as $n \to \infty$ but that $\widehat{\theta_{0.5}}$ has reduced first-order bias. The modified estimator, $\widehat{\theta_{0.5}}$, is therefore generally recommended.

Another alternative to $\widehat{\theta_{U}}$ is the _conditional_ maximum likelihood estimator, $\widehat{\theta_{C}}$. This is computed by first conditioning on ${n_{1 +}, n_{+ 1}}$. Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 +}, n_{+, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}. The maximum likelihood estimator of $\psi$, $\widehat{psi_{C}}$ is solved using iterative methods.

The conditional estimator, $\widehat{\theta_{C}}$, works better than the unconditional estimator, $\widehat{\theta_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}. Estimation of the odds ratio is made more difficult when the contingency table is _sparse_. We say that a contingency table is sparse when some of the $n_{ij}$ are small. Sparsity can occur even when the sample size, $n$, is large. Sparsity becomes more of a problem as we move from a single $2 \times 2$ table to $K$ $2 \times 2$ tables or, equivalently, a $2 \times 2 \times K$ table.

@@

Suppose we now measure the same two binary variables $X$ and $Y$ in $K$ different strata. We summarise these data in a $2 \times 2 \times K$ contingency table, $\{n_{ijk} \}$, where the $k^{th}$ 'slice' of the table corresponds to the $2 \times 2$ table for the $k^{th}$ stratum. For now, we assume that the odds ratio, $\psi = \frac{\pi_{11k} \pi_{22k}}{\pi_{12k} \pi_{21k}}$, remains constant across the $K$ tables but we allow the marginal probabilities, $\pi_{+ 1k} = 1 - \pi_{+ 2k}, \pi_{1 + k} = 1 - \pi_{2 + k}$ to vary.

We can again compute an unconditional maximum likelihood estimator, $\widehat{phi_{U}}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{phi_{0.5}}$, and a conditional maximum likelihood estimator, $\widehat{phi_{C}}$, of the log odds ratio $\theta = \log{\psi}$ \citep[see][]{Breslow:1981ta}.

__TODO: Is the unconditional maximum likelihood estimator the same as `LORCollapsed()`?__

Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi_{MH}} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{+ + k})}{n_{12k} n_{21k} / n_{+ + k}}$. \citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\theta_{MH}} = \log{\widehat{\psi_{MH}}}$, $\hat{\sigma}^{2}(\widehat{\theta_{MH}})$. This variance estimator did not appear until some 25 years after the initial publication of the Mantel-Haenszel estimator[^mh_variance]. This delay was in part due to the two asymptotic forms of $2 \times 2 \times K$ contingency tables.

@@

The first asymptotic environment, $A1$, assumes that the number of $2 \times 2$ tables, $K$, remains small while the the sample sizes in each strata, $n_{+ + k}$, grows large. The second asymptotic environment, $A2$, assumes that $K$ grows while $n_{+ + k}$ remains small. The variance estimator proposed by \citet{Robins:1986vh} is consistent under both $A1$ and $A2$. For our study of within-fragment co-methylation, $A2$ is the more relevant asymptotic environment.

@@

Under asymptotic environment $A2$, where the $2 \times 2$ tables are sparse, \citet{Liang:1985uw} developed three tests for the assumption of homogeneity of odds ratios and compared these with two tests for the same assumption but designed under $A1$. Unsurprisingly, \citeauthor{Liang:1985uw} found that the three conditional tests are more appropriate than either of the unconditional tests when the data are simulated under $A2$. The disadvantage of these conditional tests is the extra computation required to compute them. \citeauthor{Liang:1985uw} also note, however, that when both $K > 10$ and $n_{i +} > 10$, the simplest test, an unconditional score test developed under $A1$ by \citet{Zelen:1971fo} and corrected by \citet{Halperin:1977vk}, has approximately the correct size at the nominal 0.05 level.

- __TODO: Comment on the utility (or lack thereof) of homogeneity tests; i.e., rejecting the null is good evidence the null is wrong but failing to reject the null is not strong evidence that it is not wrong.__

- __TODO: See how the variance estimator behaves in some of the more pathological simulations__

@@

- the number of tables, $K$. The number of 2-tuples.
- the marginal probabilities, $\pi_{1 +} = Pr(Z_{i})$ and $\pi_{+ 1} = Pr(Z_{i'})$. Some simulations use a common set of marginal probabilities across all tables but some simulate the marginal probabilities from a specified distribution.
- the odds ratio of each table, $\psi_{k}$. Most simulations use a common odds ratio across the $K$ tables, $\psi_{k} = \psi$.
- the sample size of each table, $n_{+ + k} = d_{(i, i')}$. The number of reads containing both the $i^{th}$ and ${i'}^{th}$ methylation locus (sequencing depth). Most simulations use a truncated negative binomial distribution to model a bisulfite-sequencing experiment with mean sequencing depth of $30 \times$ but where only 2-tuples with at least $10 \times$ sequencing depth are used in downstream analyses[^nb_sim].

@@

- Use parameters $n, \pi, \psi, \theta$ and not the ones from Chapter \ref{chap:wgbs_statistical_framework} since these are more succinct.

@@

We will be studying within-fragment co-methylation by analysing a large number of sparse $2 \times 2$ contingency tables. This puts us in the realm of asymptotic environment $A2$.
Furthermore, the marginal probabilities of these tables, $\pi_{+ 1} = 1 - \pi_{+ 2}$ and $\pi_{1 +} = 1 - \pi_{2 +}$, will be very often close to the boundaries of zero and one. This is because in this case the marginal probability represents the probability that the methylation locus is methylated, and we have seen in Chapter \ref{chap:wgbs_statistical_framework} that these values, as estimated by $\beta$-values, have a bimodal distribution with each mode near the boundary.

@@
